---
title: "Clean & Reformat data: DARC to SeaTube"
output: html_document
date: "`r Sys.Date()`"
author: "Claire Hsieh with support from Sarah Groves"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up

## Load libraries

```{r, echo=FALSE}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(sf)
library(ggOceanMaps)
```

## Set user

```{r}

#user <- "Claire"
user <- "Sarah"

print_table <- FALSE
```

## Set Expendition

```{r}
expedtion <- "EX2306" # change this each time you run
```


Subset dataset based on missing values in a given column Run a conditional “if” statement based on the number of rows that either Returns a message stating no values are missing or Creates a table with the rows that are missing

```{r}

if(user == "Claire"){
  setwd("~/NOAA/NOAA.Explore.QAQC/R") 
  expedition = "EX2306_Survey_Data_20240528.csv"
  survey_df = read.csv(paste0("../../data/",expedition)) |>
    dplyr::filter(AphiaID != -999)}

if(user == "Sarah"){
  expedition = "EX2306" 
  survey_df = read.csv(paste0("C:/Users/sarah.groves/Documents/Data/EX2306/SME_clean/",expedition, "_Survey_Data_20240528.csv")) |>
    dplyr::filter(AphiaID != -999)}


# Helpful notes:
# paste0: doesn't add separators btw items else paste defaults to single space
# paste: can add optional delimiters (sep = "_")
```


# Missing Data

**Table 1.** Rows missing spatial information: Latitude
```{r Missing data Latitude, echo=FALSE}

missing_lat <- survey_df[is.na(survey_df$Latitude),]

if(nrow(missing_lat) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_lat |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_lat)
  }
  else{
    
    print(paste(missing_lat |> nrow() ,  "rows missing in latitutde column"))
   
  }
  
  
}
```

**Table 2.** Rows missing spatial information: Longitude
```{r Missing data Longitude, echo=FALSE}

missing_lon <- survey_df[is.na(survey_df$Longitude),]


if(nrow(missing_lon) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_lon |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_lon)
  }
  else{
    
  print(paste(missing_lon |> nrow() ,  "rows missing in longitude column"))
   
  }
}
```


**Table 3.** Rows missing spatial information: Depth (m)
```{r Missing data Depth , echo=FALSE}

missing_depth <- survey_df[is.na(survey_df$DepthInMeters),]

if(nrow(missing_depth) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_depth |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_depth)
  }
  else{
    print(paste(missing_depth  |> nrow() ,  "rows missing in depth column"))
   
    
  
  }
}
```
**Table 4.** Rows missing substrate information: CMECSGeoForm
```{r Missing data CMECSGeoForm, echo=FALSE}

missing_geoform <- survey_df[is.na(survey_df$CMECSGeoForm),]

if(nrow(missing_geoform) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_geoform |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_geoform)
  }
  else{
    print(paste(missing_geoform |> nrow()  ,  "rows missing in geoform column"))
   
    
  }
}
```

**Table 5.** Rows missing identification information: ScientificName
```{r Missing data ScientificName, echo=FALSE}

missing_sci_name <- survey_df[is.na(survey_df$ScientificName),]

if(nrow(missing_sci_name) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_sci_name |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_sci_name)
  }
  else{
    print(paste(missing_sci_name |> nrow()   ,  "rows missing in scientific name column"))
   
    
  }
}
```


**Table 6.** Rows missing identification information: AphiaID
```{r Missing data AphiaID, echo=FALSE}

missing_id <- survey_df[is.na(survey_df$AphiaID),]

if(nrow(missing_id) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_id |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_id)
  }
  else{
    print(paste(missing_id |> nrow()  ,  "rows missing in AphiaID column"))
   
    
  }
}
```
**Table 7.** Rows missing identification information: LifeScienceIdentifier
```{r Missing data LifeScienceIdentifier, echo=FALSE}

missing_sci_id <- survey_df[is.na(survey_df$LifeScienceIdentifier),]

if(nrow(missing_sci_id) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_sci_id |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_sci_id)
  }
  else{
    print(paste(  missing_sci_id |> nrow()   ,  "rows missing in life science identifier column"))
   
   
  }
}
```
**Table 8.** Rows missing identification information: TaxonRank
```{r Missing data TaxonRank, echo=FALSE}

missing_tax <- survey_df[is.na(survey_df$TaxonRank),]

if(nrow(missing_tax) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_tax |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_tax)
  }
  else{
    print(paste(missing_tax |> nrow() ,  "rows missing in taxon column"))
   
     
  }
}
```


**Table 9.** Rows missing identification information: DepthMethod
```{r Missing data DepthMethod, echo=FALSE}

missing_depth_method <- survey_df[is.na(survey_df$DepthMethod),]

if(nrow(missing_depth_method) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_depth_method |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_depth_method)
  }
  else{
    print(paste( missing_depth_method |> nrow()  ,  "rows missing in depth method column"))
   
    
  }
}
```
**Table 10.** Rows missing identification information: MinimumDepthInMeters
```{r Missing data MinimumDepthInMeters, echo=FALSE}

missing_min_depth <- survey_df[is.na(survey_df$MinimumDepthInMeters),]

if(nrow(missing_min_depth) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_min_depth |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_min_depth)
  }
  else{
    print(paste(missing_min_depth |> nrow() ,  "rows missing in minumum depth column"))
   
     
  }
}
```
**Table 11.** Rows missing identification information: MaximumDepthInMeters
```{r Missing data MaximumDepthInMeters, echo=FALSE}

missing_max_depth <- survey_df[is.na(survey_df$MaximumDepthInMeters),]

if(nrow(missing_max_depth) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_max_depth |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_max_depth)
  }
  else{
    print(paste(missing_max_depth |> nrow() ,  "rows missing in maximum depth column"))
   
      
  }
}
```
**Table 12.** Rows missing temporal information: ObservationDate
```{r Missing data ObservationDate, echo=FALSE}

missing_date <- survey_df[is.na(survey_df$ObservationDate),]

if(nrow(missing_date) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_date |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_date)
  }
  else{
    print(paste(missing_date |> nrow()  ,  "rows missing in date column"))
   
    
  }
}
```
**Table 13.** Rows missing temporal information: ObservationTime
```{r Missing data ObservationTime, echo=FALSE}

missing_time <- survey_df[is.na(survey_df$ObservationTime),]

if(nrow(missing_time) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_time |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_time)
  }
  else{
    print(paste(missing_time |> nrow()   ,  "rows missing in time column"))
   
    
  }
}
```
**Table 14.** Rows missing identification information: SurveyID
```{r Missing data SurveyID, echo=FALSE}

missing_survey_id <- survey_df[is.na(survey_df$SurveyID),]

if(nrow(missing_survey_id) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_survey_id |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_survey_id)
  }
  else{
    print(paste(missing_survey_id |> nrow()  ,  "rows missing in survey id column"))
   
    
  }
}
```
**Table 15.** Rows missing sensor information: Temperature
```{r Missing data Density, echo=FALSE}

missing_temp <- survey_df[is.na(survey_df$Temperature),]

if(nrow(missing_temp) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_temp |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_temp)
  }
  else{
    print(paste( missing_temp |> nrow()   ,  "rows missing in Temperature column"))
   
    
  }
}
```
**Table 16.** Rows missing sensor information: Salinity
```{r Missing data Cover, echo=FALSE}

missing_sal <- survey_df[is.na(survey_df$Salinity),]

if(nrow(missing_sal) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_sal |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_sal)
  }
  else{
    print(paste(missing_sal |> nrow()  ,  "rows missing in Salinity column"))
   
    
  }
}
```

**Table 17.** Rows missing sensor information: Oxygen
```{r Missing data Cover, echo=FALSE}

missing_o2 <- survey_df[is.na(survey_df$Oxygen),]

if(nrow(missing_o2) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_o2 |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_o2)
  }
  else{
    print(paste(missing_o2 |> nrow()  ,  "rows missing in Oxygen column"))
   
    
  }
}
```

**Table 18.** Rows missing geographical information: Habitat
```{r Missing data Habitat, echo=FALSE}

missing_habitat <- survey_df[is.na(survey_df$Habitat),]

if(nrow(missing_habitat) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_habitat |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_habitat)
  }
  else{
    print(paste(missing_habitat |> nrow()  ,  "rows missing in habitat column"))
   
    
  }
}
```
**Table 19.** Rows missing geographical information: Substrate
```{r Missing data Substrate, echo=FALSE}

missing_substrate <- survey_df[is.na(survey_df$Substrate),]

if(nrow(missing_substrate) == 0)
{
  "No missing data"
} else
{
  if (print_table){
    missing_substrate |>
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_substrate)
  }
  else{
    print(paste(missing_substrate |> nrow()  ,  "rows missing in substrate column"))
   
    
  }
}
```



# Plot Data

**Fig. 1.** Histogram of annotation Latitude.  
```{r hist of annotation Latitude, fig.height = 4, fig.width = 7, echo = FALSE}

# create unique sites
uniquesites <- unique(survey_df[, c("SurveyID", "ObservationDate", "ObservationTime", "SampleID", "Latitude" )])

Latitude_hist <- ggplot(uniquesites, aes(x=Latitude)) + 
  geom_histogram() +
  theme_bw() + 
  xlab("Latitude")

"Summary statistics of latitude:"
summary(uniquesites$Latitude)


Latitude_hist

```

**Fig. 2.** Histogram of annotation Longitude.  
```{r hist of annotation Longitude, fig.height = 4, fig.width = 7, echo = FALSE}

# create unique sites
uniquesites <- unique(survey_df[, c("SurveyID", "ObservationDate", "ObservationTime", "SampleID", "Longitude" )])

Longitude_hist <- ggplot(uniquesites, aes(x=Longitude)) + 
  geom_histogram() +
  theme_bw() + 
  xlab("Longitude")
Longitude_hist

"Summary statistics of longitude:"
summary(uniquesites$Longitude)


```

**Fig. 3.** Histogram and probability distribution of Depth.  
```{r hist of depth, fig.height = 4, fig.width = 7, echo = FALSE}

# create unique sites
uniquesites <- unique(survey_df[, c("SurveyID", "ObservationDate", "ObservationTime", "SampleID", "DepthInMeters" )]) 


depth_hist <- ggplot(uniquesites, aes(x=DepthInMeters)) + 
  geom_histogram() +
  geom_density(alpha = 0.2) +
  theme_bw() + xlab("Depth (m)")

"Summary statistics of depth (m)"
summary(uniquesites$DepthInMeters)


depth_hist

```
**Fig. 4.** Plot of ObservationDate  
```{r plot of observation date, fig.height = 4, fig.width = 7, echo = FALSE}
options(stringsAsFactors = FALSE)

# create unique sites
uniquesites <- unique(survey_df[, c("SurveyID", "ObservationDate", "ObservationTime", "SampleID" )]) 
uniquesites$ObservationDate <- as.Date(uniquesites$ObservationDate, "%m/%d/%Y",) 


date_freq <- uniquesites |>
  group_by(ObservationDate) |>
  summarise(freq = n())

obs_date_plot <- ggplot(date_freq, aes(x = ObservationDate, y = freq)) +
                        geom_bar(stat = "identity") +
                        labs(title = "Frequency of Observations by Date", x = "Observation Date", y = "Frequency") +
                        theme_minimal() +
                        scale_x_date(date_labels = "%Y-%m-%d", date_breaks = "1 day") +
                        theme(axis.text.x = element_text(angle = 45, hjust = 1))
"Summary statistics of Observation Date"
summary(uniquesites$ObservationDate)


obs_date_plot

```

**Fig. 5.** Plot of Observation Date and Time.  
```{r plot of observation time, fig.height = 4, fig.width = 7, echo = FALSE}
# create unique sites
uniquesites <- unique(survey_df[, c("SurveyID", "ObservationDate", "ObservationTime", "SampleID" )]) 
uniquesites$ObservationDate <- as.Date(uniquesites$ObservationDate, "%m/%d/%Y",) 
uniquesites$ObservationTime <- as.POSIXct(uniquesites$ObservationTime, format = "%H:%M:%S")

obs_time_plot <- ggplot(uniquesites, aes(x = ObservationDate, y = ObservationTime)) +
                        geom_point() +
                        labs(title = "Observation Date vs Time", x = "Observation Date", y = "Observation Time (Hour)") +
                        theme_minimal() +
                        scale_x_date(date_labels = "%Y-%m-%d", date_breaks = "1 day") +
                        scale_y_datetime(date_labels = "%H", date_breaks = "1 hour") +
                        theme(axis.text.x = element_text(angle = 45, hjust = 1))
"Summary statistics of Observation Time."
summary(uniquesites$ObservationTime)


obs_time_plot

### FIX Y AXIS ###
```

**Fig. 10.** Histogram and probability distribution of Temperature.  
```{r hist of temperature, fig.height = 4, fig.width = 7, echo = FALSE}

# create unique sites
uniquesites <- unique(survey_df[, c("SurveyID", "ObservationDate", "ObservationTime", "SampleID", "Temperature" )]) 


temp_hist <- ggplot(uniquesites, aes(x=Temperature)) + 
  geom_histogram() +
  geom_density(alpha = 0.2) +
  theme_bw() + xlab("Temperature (C)") + ylab("Count")

"Summary statistics of Temperature (C)"
summary(uniquesites$Temperature)


temp_hist

```

**Fig. 11.** Histogram and probability distribution of Salinity.  
```{r hist of salinity, fig.height = 4, fig.width = 7, echo = FALSE}

# create unique sites
uniquesites <- unique(survey_df[, c("SurveyID", "ObservationDate", "ObservationTime", "SampleID", "Salinity" )]) 


salinity_hist <- ggplot(uniquesites, aes(x=Salinity)) + 
  geom_histogram() +
  geom_density(alpha = 0.2) +
  theme_bw() + xlab("Salinity") + ylab("Count") # needs units

"Summary statistics of depth (m)"
summary(uniquesites$Salinity)


salinity_hist

```

**Fig. 12.** Histogram and probability distribution of Oxygen.  
```{r hist of oxygen, fig.height = 4, fig.width = 7, echo = FALSE}

# create unique sites
uniquesites <- unique(survey_df[, c("SurveyID", "ObservationDate", "ObservationTime", "SampleID", "Oxygen" )]) 


ox_hist <- ggplot(uniquesites, aes(x=Oxygen)) + 
  geom_histogram() +
  geom_density(alpha = 0.2) +
  theme_bw() + xlab("Oxygen") + ylab("Count") # needs units

"Summary statistics of depth (m)"
summary(uniquesites$Oxygen)


ox_hist
```

# Check Values

```{r}
print("CMECS Geoform")
print(unique(survey_df$CMECSGeoForm))

```


```{r}

print("Habitat")
print(unique(survey_df$Habitat))

```


```{r}

print("Substrate")
#print(unique(survey_df$Substrate))

```

# Things to add: Indivudal counts are all numeric and categorical abundance is all categorical and makes sense. 



# Plot Map - potential uupdate, color dive sites by dive number
```{r}

plt_df = unique(survey_df |> select(SurveyID, Station, Latitude, Longitude))

basemap(limits = c(-154, -130, 53, 61), bathymetry = TRUE, rotate = TRUE) + 
  ggspatial::geom_spatial_point(data = plt_df, aes(x = Longitude, y = Latitude), color = "red")
```


# Compare with OBIS Dataset - note this takes hours to run.

```{r}
library(robis)
```

```{r}
get_occurences <- check_aphiaid_near_location <- function(aphiaid) {
  # Fetch occurrence data for the given AphiaID
  occurrences <- occurrence(taxonid = aphiaid)
  return(occurrences = occurrences)
}

check_distances <- function(aphiaid, occurrences, samples_df, radius = 0.5, print_map = FALSE){
  subset_df = samples_df[samples_df$AphiaID == aphiaid, ]
  latitudes = subset_df$Latitude
  longitudes = subset_df$Longitude
  subset_df[, "Distance"] = NA
  if (nrow(occurrences) != 0) {
    print(nrow(occurrences))
    for (dist in 1:length(latitudes)){
      subset_df[dist,]$Distance = min(geosphere::distHaversine(
      matrix(c(longitudes[dist], latitudes[dist]), ncol = 2),
      matrix(c(occurrences$decimalLongitude, occurrences$decimalLatitude), ncol = 2)
    ) / 1000 )
    }
  }
  return(subset_df)
}

check_aphiaid_near_location <- function(occurences, subset_df, radius, output_dir, print_map = FALSE) {
  aphiaid = subset_df$AphiaID[1]
  num_found = sum(subset_df$Distance <= radius)
  # print(subset_df$Genus[1], subset_df$Species[1])
  print(paste("Found", num_found, "samples within", radius, "km"))
  if (num_found > 0){
    found = TRUE
    write(paste(cat(aphiaid[1]), cat(num_found[1])), paste0(output_dir, "found.csv"))
  } else {
    found = FALSE
    write(paste0(aphiaid[1], num_found[1]), file = paste0(output_dir, "checked.csv"))  }
  if (print_map) {
    map_leaflet(occurrences)
  }
  return(found)
}
```


```{r}
<- survey_df$AphiaID
samples_df = survey_df |> select(AphiaID, Genus, Species, Latitude, Longitude)
output_dir = "../data/aphiaid/"
for (i in 6:length(aphiaids)){
  radius = 0.5
  print(i)
  aphiaid = aphiaids[i]
  print(paste("Aphia ID:", aphiaid))
  occurrences <- get_occurences(aphiaid)
  # radius is in km
  subset_df <- check_distances(aphiaid, occurrences, samples_df, radius = radius)
  found <- check_aphiaid_near_location(occurrences, subset_df, radius, output_dir)
  if (found) {
    write.csv(occurrences, file = paste(output_dir, "aphiaid_",aphiaid, "_occurences.csv", sep=""))
    write.csv(subset_df, file = paste(output_dir, "aphiaid_",aphiaid, "_survey.csv", sep=""))
    write.csv(paste0(aphiaid, ""))
    # if (print_map) {
    #   map_leaflet(occurrences)
    # }
  }
}

check_aphiaid_near_location
print(results)
```

# Reformat and export data
```{r}

# Load SeaTube Template  

tmp <- read.csv("C:/Users/sarah.groves/Documents/Data/Seatube/SeaTube Template.csv")

# Create a bioogical dataframe
dat1 <- survey_df |>
  # Subset Biological observations and metadata
  dplyr::select(SurveyID, Station, Latitude, Longitude, DepthInMeters, ScientificName, Phylum, Class, Subclass, Order, Suborder, Family, Subfamily, Genus, Subgenus, Species, Morphospecies, ReporterEmail, IdentificationDate, IdentificationQualifier, ObservationDate, IndividualCount, ObservationTime, CategoricalAbundance, VerbatimSize, MinimumSize, MaximumSize, Condition, AssociatedTaxa, OccurrenceComments, LocationAccuracy) |>
  # Get rid of the ".sp" in the Scientific Name column
  dplyr::mutate(ScientificName = sub(".sp.*", "", `ScientificName`))

# Create a vector of the Taxon Rank values
TaxonRank <- c("Phylum", "Class", "Subclass", "Order", "Suborder", "Family", "Subfamily", "Genus", "Subgenus", "Species")

# Create a dataframe of the second lowest taxa for the "taxon" column
second_taxa <- dat1 |>
# Pivot to long & add TaxonRank column
  tidyr::pivot_longer(cols = c(TaxonRank), names_to = "TaxonRank", values_to = "Taxon2") |>
   dplyr::group_by(SurveyID, Station, Latitude, Longitude, DepthInMeters) |> 
  # Get rid of the lowest taxa so we can get to the 2nd lowest taxa
  dplyr::filter(ScientificName != Taxon2) |>
  # Select the last value after an NA - now the 2nd lowest taxonomic level 
  mutate(next_lowest = last(na.locf(Taxon2,fromLast=F, na.rm = T)))



# taxonomy = always "WORMS"
# parent taxon = ScientificName
# taxon = 1 level up from scientific name
# NOAA Biology/Morphospecies = Morphospecies
# creator email = ReporterEmail
# creation timestamp = IdentificationDate
# NOAA Biology/Identification Qualifier = IdentificationQualifier
# annotation timestamp = ObservationDate + ObservationTime
# NOAA Biology/Individual Count = IndividualCount
# NOAA Biology/Categorical Abundance = CategoricalAbundance
# NOAA Biology/Verbatim Size = VerbatimSize
# NOAA Biology/Minimum Size = MinimumSize
# NOAA Biology/Maximum Size = MaximumSize
# NOAA Biology/Condition = Condition 
# NOAA Biology/Associated Taxa = AssociatedTaxa
# comments = OccurenceComments
# NOAA Biology/Location Accuracy = LocationAccuracy
# NOAA Biology/Record Type = RecordType
```

