---
title: "preprocessing"
output: html_document
date: "2024-09-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up

## Load libraries

```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(sf)
```

## Set user

```{r}

#user <- "Claire"
user <- "Sarah"

```

## Set Expendition

Subset dataset based on missing values in a given column Run a conditional “if” statement based on the number of rows that either Returns a message stating no values are missing or Creates a table with the rows that are missing

# Subset dataset and immediately filter out test data (AphiaID = -999)

```{r}

if(user == "Claire"){
  setwd("~/NOAA/NOAA.Explore.QAQC/R") 
  expedition = "EX2306_Survey_Data_20240528.csv"
  survey_df = read.csv(paste0("../../data/",expedition))}

if(user == "Sarah"){
  expedition = "EX2306" 
  survey_df = read.csv(paste0("C:/Users/sarah.groves/Documents/Data/EX2306/SME_clean/",expedition, "_Survey_Data_20240528.csv")) |>
    dplyr::filter(AphiaID != -999)}


# Helpful notes:
# paste0: doesn't add separators btw items else paste defaults to single space
# paste: can add optional delimiters (sep = "_")
```

## View columns names

```{r}
# Column Options
colnames(survey_df)
```


**Table 1.** Rows missing spatial information: Latitude
```{r Missing data Latitude, echo=FALSE}

missing_lat <- survey_df[is.na(survey_df$Latitude),]

if(nrow(missing_lat) == 0)
{
  "No missing data"
} else
{
  missing_lat %>%
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_lat)  
  
}
```

**Table 2.** Rows missing spatial information: Longitude
```{r Missing data Longitude, echo=FALSE}

missing_lon <- survey_df[is.na(survey_df$Longitude),]

if(nrow(missing_lon) == 0)
{
  "No missing data"
} else
{
  missing_lon %>%
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_lon)  
  
}
```


**Table 3.** Rows missing spatial information: Depth (m)
```{r Missing data Depth , echo=FALSE}

missing_depth <- survey_df[is.na(survey_df$DepthInMeters),]

if(nrow(missing_depth) == 0)
{
  "No missing data"
} else
{
  missing_lon %>%
    dplyr::select(SurveyID, ObservationDate,ObservationTime, SampleID)
  DT::datatable(missing_depth)  
  
}
```

**Fig. 1.** Histogram of annotation Latitude.  
```{r hist of annotation Latitude, fig.height = 4, fig.width = 7, echo = FALSE}

# create unique sites
uniquesites <- unique(survey_df[, c("SurveyID", "ObservationDate", "ObservationTime", "SampleID", "Latitude" )])

Latitude_hist <- ggplot(uniquesites, aes(x=Latitude)) + 
  geom_histogram() +
  theme_bw() + 
  xlab("Latitude")

"Summary statistics of latitude:"
summary(uniquesites$Latitude)


Latitude_hist

```

**Fig. 2.** Histogram of annotation Longitude.  
```{r hist of annotation Longitude, fig.height = 4, fig.width = 7, echo = FALSE}

# create unique sites
uniquesites <- unique(survey_df[, c("SurveyID", "ObservationDate", "ObservationTime", "SampleID", "Longitude" )])

Longitude_hist <- ggplot(uniquesites, aes(x=Longitude)) + 
  geom_histogram() +
  theme_bw() + 
  xlab("Longitude")
Longitude_hist

"Summary statistics of longitude:"
summary(uniquesites$Longitude)


```

**Fig. 3.** Histogram and probability distribution of Depth.  
```{r hist of depth, fig.height = 4, fig.width = 7, echo = FALSE}

# create unique sites
uniquesites <- unique(survey_df[, c("SurveyID", "ObservationDate", "ObservationTime", "SampleID", "DepthInMeters" )]) 


depth_hist <- ggplot(uniquesites, aes(x=DepthInMeters)) + 
  geom_histogram() +
  geom_density(alpha = 0.2) +
  theme_bw() + xlab("Depth (m)")

"Summary statistics of depth (m)"
summary(uniquesites$DepthInMeters)


depth_hist

```


```{r}
### Data Cleaning ###
# filter function for missing / unexpected values


 filter_df <- function(df, col = NA, filter_type = "missing", return_type = "original", num_missing = 0, unexpected_value = -999, lat_long = NA) {
   # filter_type: 
   #   -  "missing" : filter na values
   #   - "unexpected" : filter value stored in unexpected_value
   # return_type: 
   #   -  original: return the original df filtered for specified values
   #   -  filtered: return the rows that contain specified / missing value
   # num_missing:
   #   - threshold for missing values, return subset if >  num_missing
   # unexpected_value
   #   - value to filter unexpected by
   
   if (filter_type == "missing"){
     subset_rows = is.na(df[[col]])
     if (length(subset_rows) > num_missing) {
       print(paste(length(subset_rows), "missing values in", col))
     } else {
       print(paste(nrow(filtered), "missing values in", col))
     }
    } else if (filter_type == "unexpected") {
     subset_rows = df[[col]] == unexpected_value
     num_unexp = length(subset_rows)
     print(paste(num_unexp, "unexpected values in", col))
   } else if (is.na(col) & !is.na(lat_long)){
      subset_rows = df %>%
        filter(between(Latitude, lat_long[[1]][1], lat_long[[1]][2]) &
               between(Longitude, lat_long[[2]][1], lat_long[[2]][2]))
     subset_rows <- df %>%
       mutate(in_range = between(Latitude, lat_long[[1]][1], lat_long[[1]][2]) &
                between(Longitude, lat_long[[2]][1], lat_long[[2]][2])) %>%
       pull(in_range)
     num_unexp = length(subset_rows)
     print(paste(num_unexp, "values in not in lat/long range"))
   }
   
   filtered <- df[subset_rows, ]
   df <- df[!subset_rows,]
   if (return_type == "original"){
     return(df)
   } else if (return_type == "filtered"){
     return(filtered)
   }
 }
 
# # Check missing
 column <- "TaxonRank"
 head(filter_df(survey_df, column))
 
# # Check unexpected
 column <- "AphiaID"
head(filter_df(survey_df, column, filter_type = "unexpected", return_type = "original"))

# Check latitude longitude
### Only arctic ocean for now: 
#     - latitude: [50, 70]
#     - longitude: [-140, -170]
latlong = list(c(50, 70), c(-170, -140))

# plot observations on map 
# 
head(filter_df(survey_df, column, return_type = "original", lat_long = latlong))
```

```{r}
latlong = list(c(50, 70), c(-170, -140))
head(filter_df(survey_df, column, return_type = "original", lat_long = latlong))
```

# Additional Data Cleaning:

values that dont make sense (ie -999 in AphiaID --\> Test data) Time / Location (last yr, all USA) (missing lat / long) only need to check col listed in sheet everything that has -999 need to go into annottaion platform (sea2) --\> will get data template

-   maybe learn SQL if time permits

# Plotting

-   distributions
-   geographical loc. map
-   hist. / ggmap / plotly
-   <https://r.geocompx.org/adv-map>
-   try and find some map background
-   shape files
-   plot all the diff annotations for the ROV
-   by time and by geo.

## Distribution

```{r}
 # conditions:
 ### "Temperature", "Salinity", "Oxygen", 
 # species
 ### Phylum, Class, Subclass, Order, Suborder, Family, Subfamily, Genus, Subgenus, Species, Subspecies
 # time
 ### IdentificationDate, ObservationDate, ObservationTime
 # spatial
 ### DepthInMeters, MinimumDepthInMeters, MaximumDepthInMeters
 
 
 filter_df(survey_df, col = "Temperature") %>% 
   ggplot(aes(x=Temperature)) + 
   geom_histogram()
 
 survey_df = filter_df(survey_df, col = "Temperature") 
 survey_df = filter_df(survey_df, col = "DepthInMeters") 
 survey_df = filter_df(survey_df, col = "Phylum") 

 survey_df %>% 
   ggplot(aes(x=Temperature, y=DepthInMeters, color=Phylum)) + 
   geom_point()

```

 ## Geographic Location
 
```{r}
# ### Plot geographic location using ggplot ###
# # option to color by species / depth / other condition
# # conditions:
# ### "Temperature", "Salinity", "Oxygen", 
# # species
# ### Phylum, Class, Subclass, Order, Suborder, Family, Subfamily, Genus, Subgenus, Species, Subspecies
# # time
# ### IdentificationDate, ObservationDate, ObservationTime
# # spatial
# ### DepthInMeters, MinimumDepthInMeters, MaximumDepthInMeters
# 
# 
# set.seed(3)
# survey_df = filter_df(survey_df, col="Oxygen")
# survey_sf <- survey_df %>% 
#   st_as_sf(coords = c('Longitude', 'Latitude')) %>%
#   st_set_crs(4326) # using 4326 for lat/lon decimal 
# 
# 
# ggplot() +
#   geom_sf(data = survey_sf, aes(color = Oxygens), size = 3) 

```

```{r}
### Plots geographic location + background map ###
# Makes my computer lag 

# library(sp)
# library(leaflet)
# 
# coordinates(survey_df) <- ~Longitude + Latitude
# 
# # Plot the data using leaflet
# leaflet(survey_df) %>%
#   addTiles() %>%
#   addMarkers()
```
